---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data


The variables included in this dataset are:

1. **steps**: Number of steps taking in a 5-minute interval (missing values are coded as NA ) 
2. **date**: The date on which the measurement was taken in YYYY-MM-DD format 
3. **interval**: Identifier for the 5-minute interval in which measurement was taken

Read in the data and examine the structure:

```{r, echo=T}

data = read.csv('activity.csv', header = T)
head(data)

#formats the date column as dates
data$date <- as.POSIXlt(data$date)

str(data)

```

## What is mean total number of steps taken per day?

First sum the number of steps per day:

```{r, echo=T}

stepsByDay <- tapply(data$steps, data$date, sum, na.rm=T)

```

Now create a histogram that displays the mean and median steps per day:

```{r, echo=T}

hist(stepsByDay,breaks=length(stepsByDay),main='Steps by Day',xlab = 'Steps',col='grey')

# calculate the mean and median number of steps per day
meanSteps <- mean(stepsByDay)
medianSteps <- median(stepsByDay)
        
# mark the mean and median values on the histogram
abline(v=meanSteps, lwd = 3, col = 'green')
abline(v=medianSteps, lwd = 3, col = 'blue')
        
# create legend that displays the mean and median values
legend('topright', lty = 1, lwd = 3, col = c('green','blue'),legend = 
           c(paste('Mean: ', round(meanSteps)), paste('Median: ', medianSteps)))

```

## What is the average daily activity pattern?

First average the data by time interval:

```{r, echo=T}

stepsByInt <- tapply(data$steps, data$interval, mean, na.rm=T)
head(stepsByInt)

```

Now create a time series plot of average number of steps taken by time interval:

```{r, echo=T}

plot(unlist(labels(stepsByInt)),stepsByInt, type='l',main='Average Daily Activity Pattern',xlab='Time (5 Minute Intervals)', ylab='Average Number of Steps Across All Days')

```

Finally, calculate the 5-minute time interval which averages the most number of steps.

```{r, echo=T}

# calculates the max number of steps and the interval where it occurs
maxSteps <- max(stepsByInt)
maxSteps
maxInt <- labels(stepsByInt[(which(stepsByInt==maxSteps))])
maxInt

```

## Imputing missing values

First, calculate the number of missing values in the dataset.

```{r, echo=T}
numNA <- sum(is.na(data$steps))
numNA

```

Next, fill in missing values with the average for that time period.

```{r, echo=T}
# make a copy of data
fdata <- data
head(fdata)

# function that replaces NA with the average steps for that interval
replaceNa = function(steps, int){
    needsReplaced <- is.na(steps)
    if (needsReplaced==1){
        replace <- stepsByInt[as.character(int)]
        return (replace)
    }
    return (steps)
    
}

fdata$steps <- replaceNa(fdata$steps,fdata$interval)

head(fdata)


```

Now sum the number of steps per day for the filled in data:

```{r, echo=T}

fstepsByDay <- tapply(fdata$steps, data$date, sum, na.rm=T)

```

Now create a histogram that displays the mean and median steps per day:

```{r, echo=T}

hist(stepsByDay,breaks=length(fstepsByDay),main='Steps by Day (filled data)',xlab = 'Steps',col='grey')

# calculate the mean and median number of steps per day
fmeanSteps <- mean(fstepsByDay)
fmedianSteps <- median(fstepsByDay)
        
# mark the mean and median values on the histogram
abline(v=fmeanSteps, lwd = 3, col = 'green')
abline(v=fmedianSteps, lwd = 3, col = 'blue')
        
# create legend that displays the mean and median values
legend('topright', lty = 1, lwd = 3, col = c('green','blue'),legend = 
           c(paste('Mean: ', round(fmeanSteps)), paste('Median: ',round(fmedianSteps)))

```

Finally calculate the differences between the mean and median of the original and filled data

```{r,echo=T}

meanDif <- fmeanSteps-meanSteps
meanDif
medianDif <- fmedianSteps-medianSteps
medianDif
```

So, by filling in the missing values with the average for that time interval raised the mean by 1411.959 and the median by 371.1887.

## Are there differences in activity patterns between weekdays and weekends?

```{r, echo=T}

data$weekday <- weekdays(data$date)

weekDay= function(x){
    y=character(length=length(x))
    for (i in length(x)){
        if (x[i]=="Sunday"){
            y[i]=="weekend"
        }
        else if (x[i]=="Saturday"){
            y[i]=="weekend"
        }
        else{
            y[i]=="weekday"
        }
        y[i]
    }
    return (y)
}

weekDay(data$weekday)

data$weekday
```