---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data


The variables included in this dataset are:

1. **steps**: Number of steps taking in a 5-minute interval (missing values are coded as NA ) 
2. **date**: The date on which the measurement was taken in YYYY-MM-DD format 
3. **interval**: Identifier for the 5-minute interval in which measurement was taken

Read in the data and examine the structure:

```{r, echo=T}
data = read.csv('activity.csv', header = T)
head(data)
str(data)

```

## What is mean total number of steps taken per day?

First sum the number of steps per day:

```{r, echo=T}
stepsByDay <- tapply(data$steps, data$date, sum, na.rm=T)

```

Now create a histogram that displays the mean and median steps per day:

```{r, echo=T}

hist(stepsByDay,breaks=length(stepsByDay),main = title,xlab = 'Steps',col='grey')

# calculate the mean and median number of steps per day
meanSteps <- mean(stepsByDay)
medianSteps <- median(stepsByDay)
        
# mark the mean and median values on the histogram
abline(v=meanSteps, lwd = 3, col = 'green')
abline(v=medianSteps, lwd = 3, col = 'blue')
        
# create legend that displays the mean and median values
legend('topright', lty = 1, lwd = 3, col = c('green','blue'),legend = 
           c(paste('Mean: ', meanSteps), paste('Median: ', medianSteps)))

```

## What is the average daily activity pattern?

First average the data by time interval:

```{r, echo=T}
stepsByInt <- tapply(data$steps, data$interval, mean, na.rm=T)

```

Now create a time series plot of average number of steps taken by time interval:

```{r, echo=T}
plot(data$interval,stepsByInt, type='l',main='Average Daily Activity Pattern',
    xlab='Time (5 Minute Intervals)', ylab='Average Number of Steps Across All 
    Days')

```

Finally, calculate the 5-minute time interval which averages the most number of steps.

```{r, echo=T}
max(stepsByInt)
which.max(stepsByInt)
```

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
